# Simple regression and model validation

## Data characteristics



First we observe the data characteristics:

```{r  fig.width=5, fig.height=4, message=FALSE, warning=FALSE}

The_data <- read.csv("data/Final_data.csv")
str(The_data)
summary(The_data)

```
We have here 166 different observations with 7 variables. Gender and gender are demographical data from the participants of the test. Deep, stra and surf are numerical values which are derived through means of values of single questions. Attitude is scaled to fit the same Likert scale than deep, stra and surf questions. Points describes the number of points this individual received from the test. The participants are rather young.

Lets now investigate the characteristics of the data in form of plots:

```{r echo=TRUE, fig.height=6, fig.width=8, message=FALSE, warning=FALSE}

library(ggplot2)

# plotted data
p1 <- ggplot(The_data, aes(x = attitude, y=points, col= gender))+
  geom_point()+
  geom_smooth(method = "lm")+
  ggtitle("points by gender in relation to attitude")

p2 <- ggplot(The_data, aes(x = age, y=attitude, col= gender))+
  geom_point()+
  geom_smooth(method = "lm")+
  ggtitle("Attitude by gender in relation to age")

p3 <- ggplot(The_data, aes(x = age, y = points, col = gender))+
  geom_point()+
  geom_smooth(method = "lm")+
  ggtitle("Points by gender in relation to age")

# Arrangements of the plots
ggpubr::ggarrange(
  plotlist = list(p1, p2, p3), labels = "AUTO",
  common.legend = T)

```


In the figure A you can see that in both male and female the points increase as the attitude rises. This effect however, is is stronger in males. In the figure B the attitude of males is slightly higher than in females in younger respondents. However, as we get closer to the 50 years of age, the number of observations decrease and the separation is not so clear anymore. Majority of the respondents are 20-35 years old. In the figure C you can see that the points of older men are clearly lower than younger men. This however is not observed with females. 

## Regression model

Then the regression model of the data:

```{r echo=TRUE, fig.height=3, fig.width=4, message=FALSE, warning=FALSE}

library(GGally)
library(ggplot2)

my_model <- lm(points ~ attitude + deep + stra + surf, data = The_data)
summary(my_model)

```
Based on the coefficients, the deep, stra and surf values does not statistically significantly predict the increase or decrease of the points in the test. However, attitude can predict the increase on test points (if none of the other variables are changed), as the p-value of the test is 1.34e-08. According to the model, one unit increase in attitude predict a 3.5 +/- 0.6 points better outcome than the latter attitude unit. However, as the residuals between the real value and the predicted value are quite high (maximum of 11.6, minimum of - 16.7 and a median of 0.79) the result should be interpreted within certain limits. This is further showed with a the R-squared value, which indeed is quite big (0.21). 

Let's observe this test data with figures. However, to make this more clear let's make a new model where the deep, stra and surf variables are removed since the attitude data was the only significant variable in the model.

```{r echo=TRUE, fig.height=3, fig.width=4, message=FALSE, warning=FALSE}
# simpler model
my_model1 <- lm(points ~ attitude, data = The_data)

#the residuals vs. fitted values, normal QQ, and residuals vs leverage plots
par(mar = c(4, 4, .1, .1))
plot(my_model1, which = c(1,2,5))

```
The linear model assumes linearity and a normally distributed residual. The model seems to meet the normality assumption, evident from the QQ-plot. Additionally, there's consistent variance in errors, shown in the residual plot, and no points significantly impacting the model. In conclusion, the model effectively describes the relationship between the points and attitude.

That's all folks!




